---
title: "rbtt"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(reshape)
library(rbtt)
library(microbenchmark)
library(parallel)

options(digits = 4)
```

# rbtt (Robust bootstrap-based $t$-test)

[![CRAN version](https://www.r-pkg.org/badges/version/rbtt)](https://cran.r-project.org/package=rbtt) 
[![Travis-CI Build Status](http://travis-ci.org/WannabeSmith/rbtt.svg?branch=master)](http://travis-ci.org/WannabeSmith/rbtt) 
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/WannabeSmith/rbtt?branch=master&svg=true)](https://ci.appveyor.com/project/WannabeSmith/rbtt)

## Overview

**rbtt is an alternative bootstrap-based $t$-test aiming to reduce type-I error for non-negative, zero-inflated data**

Tu & Zhou (1999) showed that comparing the means of populations whose data-generating distributions are non-negative with excess zero observations is a problem of great importance in the analysis of medical cost data. In the same study, Tu & Zhou discuss that it can be difficult to control type-I error rates of general-purpose statistical tests for comparing the means of these particular data sets. This package allows users to perform a modified bootstrap-based t-test that aims to better control type-I error rates in these situations.

## Usage

Let's say we have some non-negative data with clumping at zero:

```{r}
x <- rbinom(50, 1, 0.5) * rlnorm(50, 0, 1)
y <- rbinom(150, 1, 0.3) * rlnorm(150, 2, 1)
```

```{r, echo=FALSE, warning=FALSE}
names(x) <- rep("x", length(x))
names(y) <- rep("y", length(y))
data <- c(x, y)
ggdf <- data.frame("value" = data, "variable" = names(data))

ggplot(ggdf, aes(x = value, fill = variable)) +
  geom_density(alpha = 0.5, position = "identity") + xlim(0, 30) + 
  ggtitle("Densities of x and y (truncated plot at value = 30)")
```

Then we may compute rbtt-based $t$-tests to compare the means:

```{r}
# Use ‘method = 1’ for a two-sample, two-sided rbtt under the equal variance assumption,
rbtt(x, y, n.boot=999, method = 1)

# Use ’method = 2' for a two-sample, one-sided rbtt without the equal variance assumption
rbtt(x, y, n.boot=999, method = 2)
```

Alternatively, you can specify ```method = "both"``` to perform both methods simultaneously (this is also done by default).

## Parallelize rbtt

```{r}
# Compare speed when using single-core versus multiple-core rbtt on 99999 bootstrap resamples
system.time(rbtt(x, y, n.boot = 99999, method = 1, n.cores = 1))

system.time(rbtt(x, y, n.boot = 99999, method = 1, n.cores = 3))
```

## Comparison between rbtt and t.test

First, we perform some simulations.
```{r}
n.sim <- 999

t.test.results <- numeric(n.sim)
rbtt.results <- numeric(n.sim)

pval.table.list <- mclapply(1:n.sim, function(i)
{
  # True means are equal
  x <- rbinom(50, 1, 0.5) * rlnorm(50, 1.15, 1)
  y <- rbinom(150, 1, 0.5) * rlnorm(150, 1.15, 1)
  
  t.test.result <- t.test(x, y)$p.value
  rbtt.result <- rbtt(x, y, n.boot = 999, method = 1)$p.value
  
  return(c(t.test.result, rbtt.result))
}, mc.cores = 4)

pval.table <- do.call(rbind, pval.table.list)
```

Now, let's evaluate the type-I of these simulations using a significance level of 0.05.
```{r}
# t.test type-I error with significance level of 0.05:
sum(pval.table[,1] < 0.05) / n.sim

# rbtt type-I error with significance level of 0.05
sum(pval.table[,2] < 0.05) / n.sim
```

More accurate p-values and type-I error estimates can be obtained by increasing ```n.boot``` and ```n.sim```, respectively
